Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2018-07-12 paper.tex
(+) Overview of Ceu

\usepackage{cite}

@@

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    @const, @pure, @safe, CHAN, FOREVER, PAR, PROC, SIGNAL, abort, and,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, s, signal, spawn, tag, then, this, traverse, until,
    code, public, private,
    var, watching, when, with},
}
\lstset{
  basicstyle=\footnotesize\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\rmfamily\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  mathescape=true,
  numbersep=4pt,
  numberstyle=\tiny,
  %upquote=true,
}

@@

In this work, we advocate structured synchronous reactive programming as a more
productive alternative for game logic development.
We present a qualitative case study of rewriting Pingus from C++ to \CEU.

\CEU~\cite{ceu.sensys13,ceu.mod15} is a Esterel-based~\cite{esterel.ieee91}
programming language that originally targets embedded soft real-time systems.
It aims to offer a concurrent, safe, and expressive alternative to C with the
characteristics that follow:
%
\begin{itemize}
\item \textbf{Reactive:} code only executes in reactions to events.
\item \textbf{Structured:} programs use structured control mechanisms, such as
    \code{spawn} and \code{await} (to create and suspend activities).
\item \textbf{Synchronous:} reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{itemize}
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage.
The runtime is single threaded and does not rely on garbage collection for
memory management~\cite{ceu.sensys13}.
%
Existing work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\%)~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.

@@

The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU and describes our approach to identify and rewrite the control flow in the
game.
Section~\ref{sec.pats} discusses four case studies in detail which are
categorized in recurrent control-flow patterns.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.

@@

The rewriting process consisted of identifying sets of callbacks in C++
implementing control flow in the game and translating them to \CEU using
appropriate structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
As a general rewriting rule, we identify control-flow behaviors in the C++
codebase by looking for class state members with identifiers resembling verbs,
statuses, and counters (e.g.,
\code{pressed},
\code{particle\_thrown},
\code{mode}, and
\code{delay\_count}).
Good chances are that such variables encode some form of control-flow
progression that cross multiple callback invocations.
%
Not all state follows these conventions, but they help finding classes that are
heavy on control flow quickly at the beginning of the process.

@@

%\begin{figure}
%\centering
%\includegraphics[width=\columnwidth]{double-click-opt}
%\caption{Double click detection.
%\label{fig.armageddon}
%}
%\end{figure}
%(Figure~\ref{fig.armageddon}).

@@

\begin{figure}[t]
%\begin{minipage}[h]{0.50\linewidth}
\begin{lstlisting}[numbers=right]
ArmageddonButton::ArmageddonButton(<...>):
    RectComponent(<...>),
    pressed(false); // button is not initially pressed
    press_time(0);  // how long since 1st click?
    <...>
{
    <...>
}

@@

void ArmageddonButton::on_click (<...>) {
    if (pressed) {
      send_armageddon_event();
    } else {
      pressed = true;
    }
}
\end{lstlisting}
%\centering\small{\ax Implementation in C++}
%\end{minipage}%
%
%\begin{minipage}[h]{0.50\linewidth}
%\begin{lstlisting}
\end{figure}
\begin{figure}[t]
\begin{lstlisting}[xleftmargin=2em]
do
    var RectComponent but = <...>;
    <...>
    loop do
      await but.on_click;
      watching 1s do
        await but.on_click;
        break;
      end
    end
    <...>
    emit game.armageddon;
end
\end{lstlisting}
%\centering\small{\bx Implementation in \CEU}
%\end{minipage}%
%\rule{8.4cm}{0.37pt}
\caption{ Detecting double-clicks in the \emph{Armageddon button}.
\label{lst.armageddon}
}
\end{figure}

@@

Figure~\ref{lst.armageddon}.a shows the C++ implementation for the class
\code{ArmageddonButton} with methods for rendering the button and handling
mouse and timer events.
The code in the figure focus on the double click detection and hides unrelated
parts with \code{<...>}.
%
The methods \code{update} (ln 14--26) and \code{on\_click} (ln 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the base class
\code{RectComponent} (ln 2), while the callback \code{update} continuously
reacts to the passage of time, frame by frame.
%Callbacks are short lived because they must react to input as fast as possible
%to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} (ln 3) to track the
first click (ln 32).
It also initializes the variable \code{press\_time} (ln 4) to count the time
since the first click (ln 16--17).
If another click occurs within 1 second, the class signals the double click to
the application (ln 29--30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln 18--21).
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variable \code{pressed}, and the arrows 
represent the callbacks manipulating it.
%
Note in Figure~\ref{lst.armageddon}.a how the accesses to the state variables
are spread across the entire class: the distance between the initialization of
\code{pressed} (ln  3) and the last access to it (ln 32) is over 40 lines in
the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}, which is defined in middle of the class (ln 10--12), can
potentially access them.

\CEU supports structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
In Figure~\ref{lst.armageddon}.b, the loop to detect double clicks (ln 4--10)
awaits the first click (ln 5) and then, while watching 1 second (ln 6--9),
awaits the second click (ln 7).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln 8) and the \code{emit} in sequence signals the double click to the
application (ln 12).
Otherwise, the \code{watching} block as a whole aborts after 1 second  and the
loop restarts, falling back to the first click \code{await} (ln 5).
%
Double click detection in \CEU does not rely on state variables and is entirely
self-contained in the \code{loop} body.
Also, those 7 lines of code \emph{only} detect the double click, leaving the
actual effect (ln 12) as well as all unrelated code (such as redrawing the
button) to happen outside the loop.

@@

void Bomber::draw (SceneContext& gc) {
    // 3. particles, terrain, explosion sprite
    // 4. tick: hides the explosion sprite
    if (spr.frame()==13 && !gfx_ok) {
      gfx_ok = true;
      gc.color().draw(explo_surf, <...>);
    }
    gc.color().draw(spr, pingu->get_pos());
}
%\end{lstlisting}
%\centering\small{\ax Implementation in C++}
%\end{minipage}
\end{figure}
%
\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[xleftmargin=2em]
code/await Bomber (void) -> ActionName
do
    <...>
    spawn Mover(); // movement in background
    var Sprite spr = spawn Sprite(<...>);
                 // frame animation in background
    watching spr do
      // 1. plays a "Oh no!" sound.
      {play_sound("ohno")};

@@

In C++, the class \code{Bomber} in Figure~\ref{lst.bomber}.a defines the
callbacks \code{draw} and \code{update} to manage the state machine of
Figure~\ref{fig.bomber.fsm}.
%
The class first defines one state variable for each effect to perform
(ln 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln 11).
The \code{update} callback (ln 14--38) first updates the pingu animation and
movement on every frame, regardless of its current state (ln 15--16).
When the animation reaches the 10th frame, it switches to \emph{state-2} and
plays the ``Bomb!'' sound (ln 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln 24--32 and
41--46).
The explosion sprite appears in a single frame in \emph{state-4} (ln 45).
Finally, the pingu dies after the animation frames terminate (ln 34--37).
%
Note that a single numeric state variable would suffice to track the states as
in Figure~\ref{fig.bomber.fsm}, but the original developers probably chose to
encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., methods with mutable state).
%
Like the double click detection in C++, note that the state machine is encoded
across 3 different methods, each intermixing code with unrelated functionality
(e.g., changing frames, moving, and redrawing).

The equivalent code in \CEU for the \emph{Bomber action} in
Figure~\ref{lst.bomber}.b does not rely on state variables and reflects the
sequential state machine implicitly, using \code{await} statements to separate
the effects in direct style.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
retains runtime state, such as local variables and the program counter, across
reactions to events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other
\code{code/await} abstractions and execute in separate through the \code{spawn}
primitive (ln 4--5).
%In \CEU, if multiple abstractions react to the same event, the scheduler
%employs lexical order to preserve determinism, i.e., the \code{spawn} that
%appears first in the source code reacts first~\cite{ceu.mod15}.
The event \code{game.update} (ln 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every game frame.
%
The code tracks the animation aliveness (ln 7--27) and, on termination,
performs the last bomber effect, killing the pingu (ln 30).
As soon as the animation starts, the code performs the first effect (ln 9).
The intermediate effects are performed when the corresponding conditions occur
(ln 12,16,24).
The \code{do-end} block (ln 19--25), restricts the lifespan of the
single-frame explosion sprite (ln 21): after the next game tick (ln 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all effects occur in a contiguous
chunk of code (ln 7--30), which handles no extra functionality.
\end{comment}

@@

Among all 65 implementation files in \CEU, we found 29 cases in 25 files that
use structured mechanisms to substitute states machines.
They typically manifest as \code{await} statements in sequence (e.g.,
ln 5,7 in Figure~\ref{lst.armageddon} and ln 12,16,24 in
Figure~\ref{lst.bomber}).
%or in aborting constructs such as \code{par/or} and \code{watching}.

@@

\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=right]
StoryDot::StoryDot(FileReader& reader) :
    show_credits(false), // do not show by default
{
    <...>
    reader.read("credits", show_credits);
}             // from file

@@

void StoryScreenComp::next_text() {
    if (!displayed) {
      <...>
    } else {
      <...>
      if (!pages.empty()) {
        <...>
      } else {
        if (show_credits) {
          replace(<Credits>(<...>));
        } else {
          pop();
        }
      }
    }
}
%\end{lstlisting}
%\centering\small{\ax Implementation in C++}
%\end{minipage}
\end{figure}
%
\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[xleftmargin=2em]
loop do
    var int ret = await Worldmap();
    if ret==CREDITS or ret==BACK then
      <...>
      var bool is_click = await Story();
      if is_click and ret==CREDITS then
        <...>
        await Credits();
      end
    else
      <...>
    end
end
\end{lstlisting}
%\centering\small{\bx Implementation in \CEU}
%\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Transition from \emph{Story} to \emph{Credits screen}.
\label{lst.story}
}
\end{figure}

@@

In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln 1--12) first
reads the level file (ln 5) to check whether it is a closing story and should,
after termination, show the \emph{Credits screen}.
%
The boolean variable \code{show\_credits} (ln 2,5,10) is passed to the class
\code{StoryScreen} (ln 10) and represents the screen continuation, i.e., what
to do after showing the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln 16--40).
%
When the method \code{next\_text} has no story pages left to display
(ln 32--38), it decides where to go next, depending on the continuation flag
\code{show\_credits} (ln 33).

In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
the screens to show as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln 2), which shows the map and lets
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln 4--9), we invoke the \code{Story}
and await its termination (ln 5).
After showing the story, we check the returned values (ln 6) to perhaps show
the \code{Credits} screen (ln 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ uses an explicit stack to push the \code{Worldmap} screen (not
          shown in Figure~\ref{lst.story}.a).
    \item \CEU invokes the \code{Worldmap} screen expecting a return value
          (Figure~\ref{lst.story}.b, ln 2).
    \end{itemize}
\item \code{Worldmap} (\emph{blue dot click}) $\longrightarrow$ \code{Story}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pushes the \code{Story} screen passing the continuation flag
          (Figure~\ref{lst.story}.a, ln 10).
    \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
          (Figure~\ref{lst.story}.b, ln 2,5).
    \end{itemize}
\item \code{Story} $\longrightarrow$ \code{Credits}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ replaces the current \code{Story} screen with the \code{Credits}
          screen (Figure~\ref{lst.story}.a, ln 34).
    \item \CEU invokes the \code{Credits} screen after the \code{await Story}
          returns (Figure~\ref{lst.story}.b, ln 8).
    \end{itemize}
\item \code{Credits} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
          screen (not shown in Figure~\ref{lst.story}.a).
    \item \CEU uses an enclosing \code{loop} to restart the process
          (Figure~\ref{lst.story}.b, ln 1--13).
    \end{itemize}
\end{enumerate}

>>>>

%\usepackage{cite}
\usepackage[noadjust]{cite}
\renewcommand{\citepunct}{,\penalty\citepunctpenalty\,}
\renewcommand{\citedash}{--}% optionally

@@

\usepackage{listings}
\lstdefinelanguage{ceu}{%
  language=C,
  morekeywords={%
    @const, @pure, @safe, NEVER, FOREVER, PAR, PROC, SIGNAL, abort, and,
    await, bool, call, class, data, define, deterministic, do, each, else,
    emit, end, escape, event, every, finalize, hor, implementation, in,
    input, interface, loop, min, native, new, nohold, not, or, output, par,
    pool, pure, return, s, h, ms, signal, spawn, tag, then, this, traverse, until,
    code, public, private,
    var, watching, when, with},
}
\lstset{
  basicstyle=\footnotesize\ttfamily,
  captionpos=b,
  columns=flexible,
  commentstyle=\rmfamily\itshape,
  escapeinside={||},
  frame=tb,
  keepspaces=true,
  keywordstyle=\bfseries,
  language=ceu,
  mathescape=true,
  numbersep=4pt,
  numberstyle=\tiny,
  %xleftmargin=0.5cm,
  %xrightmargin=0.5cm,
  %upquote=true,
}

@@

In this work, we advocate structured synchronous reactive programming (SSRP) as
a more productive alternative for game logic development.
We present a qualitative case study of rewriting Pingus from C++ to \CEU.
%
\CEU~\cite{ceu.tecs17} is a Esterel-based~\cite{esterel.ieee91}
programming language that originally targets embedded soft real-time systems.
It aims to offer a concurrent, safe, and expressive alternative to C.
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage.
The runtime is single threaded and does not rely on garbage collection for
memory management~\cite{ceu.sensys13}.
%
Existing work in the context of embedded sensor networks evaluates the
expressiveness of \CEU in comparison to event-driven code in C and attests a
reduction in source code size (around 25\%) with a small increase in memory
usage (around 5--10\%)~\cite{ceu.sensys13}.
%
\CEU has also been used in the context of multimedia
systems~\cite{ceumedia.webmedia16} and games~\cite{ceu.mod15}.

@@

\begin{comment}
The rest of the paper is organized as follows:
Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
\CEU and describes our approach to identify and rewrite the control flow in the
game.
Section~\ref{sec.pats} discusses four case studies in detail which are
categorized in recurrent control-flow patterns.
Section~\ref{sec.related} discusses related work.
Section~\ref{sec.conclusion} concludes the paper.
\end{comment}

\section{An Overview of \CEU}
\label{sec.ceu}

\CEU is a synchronous reactive language in which programs evolve in a sequence
of discrete reactions to external events with the characteristics that follow:
%
\begin{itemize}
\item \textbf{Reactive:} code only executes in reactions to events.
\item \textbf{Structured:} programs use structured control mechanisms, such as
    \code{spawn} and \code{await} (to create and suspend activities).
\item \textbf{Synchronous:} reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{itemize}
%
\CEU is designed for control-intensive applications, supporting concurrent
lines of execution, known as \emph{trails}, and instantaneous broadcast
communication through events.
%
Computations within a reaction (such as expressions, assignments, and system
calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment.

\begin{lstlisting}[
  float=t,
  caption={A compact reference of \CEU.},
  label={lst.syntax},
]
// Declarations
input <type> <id>   // declares an external input event
event <type> <id>   // declares an internal event
var   <type> <id>   // declares a variable

// Event handling
<id> = await <id>   // awaits event and assigns received value
emit <id>(<exp>)    // emits event passing a value

// Control flow
<stmt> ; <stmt>                         // sequence
if <exp> then <stmt> else <stmt> end    // conditional
loop do <stmt> end                      // repetition
do <stmt> end                           // explicit block

// Logical parallelism
par/or  do <stmt> with <stmt> end   // terminates with any
par/and do <stmt> with <stmt> end   // terminates with all

// Abstractions
code <id> (<pars>) -> <type> do <stmt> end
<id> = await <id>(<args>)       // executes and awaits
spawn <id>(<args>) [in <id>]    // executes and continues
pool[<num>] <id> <id>           // holds instances

// Assignment & Integration with C/C++
<id> = <exp>            // assigns a value to a variable
_<id>(<args>)           // calls C/C++ (id starts with `_')
\end{lstlisting}

\begin{lstlisting}[
  numbers=right,xrightmargin=0.5cm,
  float=t,
  caption={An example in \CEU that blinks an LED every 500ms in the background
          and awaits a button click or 1 hour to terminate.},
  label={lst.ceu},
]
input  void BUTTON;
output bool LED;

code Blink (var int period) -> NEVER do
    loop do
        emit LED(true);
        await (period) ms;
        emit LED(false);
        await (period) ms;
    end
end

spawn Blink(500);
par/or do
    await BUTTON;
with
    await 1h;
end
\end{lstlisting}

Listing~\ref{lst.syntax} shows a compact reference of \CEU and
Listing~\ref{lst.ceu} shows a simple example that blinks an LED and terminates
on a button press or after 1 hour of execution.
%
The example declares two external events (ln~1--2): an input that represents
a button and an output that represents an LED.
%
The blink behavior uses a \code{code} abstraction (ln~4--11) which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}, but which can also
interact directly with the environment through \code{await} statements
(ln.~7,9).
%
The \code{spawn} statement (ln~13) instantiates a \code{Blink} abstraction to
execute in the background and the program then creates two trails to await
two events at the same time (ln~14--18).
Since the composition is a \code{par/or}, it will terminate when either of the
two events occur proceeding to the end of the file and consequently also
terminating the program.

In \CEU, every execution path within loops must contain at least one
\code{await} statement to an external input
event~\cite{ceu.sensys13,esterel.primer}.
%
On the one hand, \CEU shares a restriction with event-driven programming:
computations that take a non-negligible time to run (e.g., cryptography or
image processing algorithms) violate the zero-delay hypothesis, and thus cannot
be directly implemented.
%
On the other hand, all reactions to the external environment are guaranteed to
be computed in bounded time, ensuring that games progress with
time~\cite{ceu.lctes18}.

@@

The rewriting process consisted of identifying sets of callbacks in C++
implementing control flow in the game and translating them to \CEU using
appropriate structured constructs.
%
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.
%
As a general rewriting rule, we identify control-flow behaviors in the C++
codebase by looking for class state members with identifiers resembling verbs,
statuses, and counters (e.g.,
\code{pressed},
\code{particle\_thrown},
\code{mode}, and
\code{delay\_count}).
Good chances are that such variables encode some form of control-flow
progression that cross multiple callback invocations.
%
Not all state follows these conventions, but they helped finding classes that
are heavy on control flow quickly at the beginning of the process.

@@

@@

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={C++: Detecting double-clicks in the \emph{Armageddon button}.},
  label={lst.armageddon.a},
]
ArmageddonButton::ArmageddonButton(<...>):
    RectComponent(<...>),
    pressed(false); // button is not initially pressed
    press_time(0);  // how long since 1st click?
    <...>
{
    <...>
}

@@

void ArmageddonButton::on_click (<...>) {
    if (pressed) {
      send_armageddon_event();
    } else {
      pressed = true;
    }
}
\end{lstlisting}

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={\CEU: Detecting double-clicks in the \emph{Armageddon button}.},
  label={lst.armageddon.b},
]
do
    var RectComponent but = <...>;
    <...>
    loop do
      await but.on_click;
      par/or do
        await 1s;
      with
        await but.on_click;
        break;
      end
    end
    <...>
    emit game.armageddon;
end
\end{lstlisting}

@@

Listing~\ref{lst.armageddon.a} shows the C++ implementation for the class
\code{ArmageddonButton} with methods for rendering the button and handling
mouse and timer events.
The code in the figure focus on the double click detection and hides unrelated
parts with \code{<...>}.
%
The methods \code{update} (ln 14--26) and \code{on\_click} (ln 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the base class
\code{RectComponent} (ln 2), while the callback \code{update} continuously
reacts to the passage of time, frame by frame.
%Callbacks are short lived because they must react to input as fast as possible
%to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} (ln 3) to track the
first click (ln 32).
It also initializes the variable \code{press\_time} (ln 4) to count the time
since the first click (ln 16--17).
If another click occurs within 1 second, the class signals the double click to
the application (ln 29--30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln 18--21).
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variable \code{pressed}, and the arrows 
represent the callbacks manipulating it.
%
Note in Listing~\ref{lst.armageddon.a} how the accesses to the state variables
are spread across the entire class: the distance between the initialization of
\code{pressed} (ln  3) and the last access to it (ln 32) is over 40 lines in
the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}, which is defined in middle of the class (ln 10--12), can
potentially access them.

\CEU supports structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
In Listing~\ref{lst.armageddon.b}, the loop to detect double clicks (ln 4--12)
awaits the first click (ln 5) and then, awaits either a 1-second timeout (ln 7)
or the second click (ln 9).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln 10) and the \code{emit} in sequence signals the double click to the
application (ln 14).
Otherwise, the \code{par/or} block as a whole aborts after 1 second  and the
loop restarts, falling back to the first click \code{await} (ln 5).
%
Double click detection in \CEU does not rely on state variables and is entirely
self-contained in the \code{loop} body.
Also, those 7 lines of code \emph{only} detect the double click, leaving the
actual effect (ln 12) as well as all unrelated code (such as redrawing the
button) to happen outside the loop.

@@

void Bomber::draw (SceneContext& gc) {
    // 3. particles, terrain, explosion sprite
    // 4. tick: hides the explosion sprite
    if (spr.frame()==13 && !gfx_ok) {
      gfx_ok = true;
      gc.color().draw(explo_surf, <...>);
    }
    gc.color().draw(spr, pingu->get_pos());
}
%\end{lstlisting}
%\centering\small{\ax Implementation in C++}
%\end{minipage}
\end{figure}
%
\begin{figure}[t]
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[xleftmargin=2em]
code Bomber (void) -> ActionName
do
    <...>
    spawn Mover(); // movement in background
    var Sprite spr = spawn Sprite(<...>);
                 // frame animation in background
    watching spr do
      // 1. plays a "Oh no!" sound.
      {play_sound("ohno")};

@@

In C++, the class \code{Bomber} in Listing~\ref{lst.bomber.a} defines the
callbacks \code{draw} and \code{update} to manage the state machine of
Figure~\ref{fig.bomber.fsm}.
%
The class first defines one state variable for each effect to perform
(ln 4--7).
The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
(ln 11).
The \code{update} callback (ln 14--38) first updates the pingu animation and
movement on every frame, regardless of its current state (ln 15--16).
When the animation reaches the 10th frame, it switches to \emph{state-2} and
plays the ``Bomb!'' sound (ln 18--22).
The state variable \code{sound\_played} is required because the sprite frame
doesn't necessarily advance on every \code{update} invocation (e.g.,
\code{update} may execute twice during the 10th frame).
The same reasoning and technique applies to \emph{state-3} (ln 24--32 and
41--46).
The explosion sprite appears in a single frame in \emph{state-4} (ln 45).
Finally, the pingu dies after the animation frames terminate (ln 34--37).
%
Note that a single numeric state variable would suffice to track the states as
in Figure~\ref{fig.bomber.fsm}, but the original developers probably chose to
encode each state in an independent boolean 
variable to rearrange and experiment with them during development.
Still, due to the short-lived nature of callbacks, state variables are 
unavoidable and are actually the essence of object-oriented programming
(i.e., methods with mutable state).
%
Like the double click detection in C++, note that the state machine is encoded
across 3 different methods, each intermixing code with unrelated functionality
(e.g., changing frames, moving, and redrawing).

The equivalent code in \CEU for the \emph{Bomber action} in
Listing~\ref{lst.bomber.b} does not rely on state variables and reflects the
sequential state machine implicitly, using \code{await} statements to separate
the effects in direct style.
%
The \code{Bomber} is a \code{code} abstraction of \CEU, which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
retains runtime state, such as local variables and the program counter, across
reactions to events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other
\code{code} abstractions and execute in separate through the \code{spawn}
primitive (ln 4--5).
%In \CEU, if multiple abstractions react to the same event, the scheduler
%employs lexical order to preserve determinism, i.e., the \code{spawn} that
%appears first in the source code reacts first~\cite{ceu.mod15}.
The event \code{game.update} (ln 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every game frame.
%
The code tracks the animation aliveness (ln 7--27) and, on termination,
performs the last bomber effect, killing the pingu (ln 30).
As soon as the animation starts, the code performs the first effect (ln 9).
The intermediate effects are performed when the corresponding conditions occur
(ln 12,16,24).
The \code{do-end} block (ln 19--25), restricts the lifespan of the
single-frame explosion sprite (ln 21): after the next game tick (ln 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all effects occur in a contiguous
chunk of code (ln 7--30), which handles no extra functionality.
\end{comment}

@@

Among all 65 implementation files in \CEU, we found 29 cases in 25 files that
use structured mechanisms to substitute states machines.
They typically manifest as \code{await} statements in sequence (e.g.,
ln 5,7 in Listing~\ref{lst.armageddon.b}).
%or in aborting constructs such as \code{par/or} and \code{watching}.

@@

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={C++: Transition from \emph{Story} to \emph{Credits screen}.},
  label={lst.story.a},
]
StoryDot::StoryDot(FileReader& reader) :
    show_credits(false), // do not show by default
{
    <...>
    reader.read("credits", show_credits);
}             // from file

@@

void StoryScreenComp::next_text() {
    if (!displayed) {
      <...>
    } else {
      <...>
      if (!pages.empty()) {
        <...>
      } else {
        if (show_credits) {
          replace(<Credits>(<...>));
        } else {
          pop();
        }
      }
    }
}
\end{lstlisting}

\begin{lstlisting}[
  numbers=left,xleftmargin=0.5cm,
  float=t,
  caption={\CEU: Transition from \emph{Story} to \emph{Credits screen}.},
  label={lst.story.b},
]
loop do
    var int ret = await Worldmap();
    if ret==CREDITS or ret==BACK then
      <...>
      var bool is_click = await Story();
      if is_click and ret==CREDITS then
        <...>
        await Credits();
      end
    else
      <...>
    end
end
\end{lstlisting}

@@

In C++, the class \code{StoryDot} in Listing~\ref{lst.story.a} (ln 1--12) first
reads the level file (ln 5) to check whether it is a closing story and should,
after termination, show the \emph{Credits screen}.
%
The boolean variable \code{show\_credits} (ln 2,5,10) is passed to the class
\code{StoryScreen} (ln 10) and represents the screen continuation, i.e., what
to do after showing the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln 16--40).
%
When the method \code{next\_text} has no story pages left to display
(ln 32--38), it decides where to go next, depending on the continuation flag
\code{show\_credits} (ln 33).

In \CEU, the \code{loop} of Listing~\ref{lst.story.b} controls the flow between
the screens to show as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln 2), which shows the map and lets
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln 4--9), we invoke the \code{Story}
and await its termination (ln 5).
After showing the story, we check the returned values (ln 6) to perhaps show
the \code{Credits} screen (ln 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ uses an explicit stack to push the \code{Worldmap} screen (not
          shown in Listing~\ref{lst.story.a}).
    \item \CEU invokes the \code{Worldmap} screen expecting a return value
          (Listing~\ref{lst.story.b}, ln 2).
    \end{itemize}
\item \code{Worldmap} (\emph{blue dot click}) $\longrightarrow$ \code{Story}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pushes the \code{Story} screen passing the continuation flag
          (Listing~\ref{lst.story.a}, ln 10).
    \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
          (Listing~\ref{lst.story.b}, ln 2,5).
    \end{itemize}
\item \code{Story} $\longrightarrow$ \code{Credits}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ replaces the current \code{Story} screen with the \code{Credits}
          screen (Listing~\ref{lst.story.a}, ln 34).
    \item \CEU invokes the \code{Credits} screen after the \code{await Story}
          returns (Listing~\ref{lst.story.b}, ln 8).
    \end{itemize}
\item \code{Credits} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
          screen (not shown in Listing~\ref{lst.story.a}).
    \item \CEU uses an enclosing \code{loop} to restart the process
          (Listing~\ref{lst.story.b}, ln 1--13).
    \end{itemize}
\end{enumerate}
